JEThread: A Thread Pool Implementation in C

### 2018-05-04 thread-pool-implementation/paper.tex
done up to internal sync

% Define document title and author
	\title{Thread Pool Implementation in C}
	\author{Jialin Li and Edward Hu}
	\maketitle

% Write abstract here
\begin{abstract}
	This is a short report on our thread pool performance.
\end{abstract}

% Each section begins with a \section{title} command
\section{Introduction}
	% \PARstart{}{} creates a tall first letter for this first paragraph
	\PARstart{T}{hread} is a basic unit of CPU utilization, it consists of program counter, register set, and its own stack. Compared to processes, threads are much more lighweight since virtual memory space remains the same during a thread switch, which is not true for process switch. However, even threads sometimes pose overhead that will affect the performance of the program. Initializing and cleaning threads can pose a certain amount of overhead. For an parallel application that creates and destroys a large number of threads that each only runs for a short time, the thread management mechanism can create unnecessary resource waste. Thread pool can be useful in this type of scenario as it reduces the complexity of thread management and the overhead involved in thread creation and cleaning. The thread pool in implement in \textbf{GNU C}.

% Main Part
\section{Motivation}
	We want to be able to implement a thread pool that is convinient to use. We aim to create a provide an interface that is easy for programmers to use. Thus we minimize the number of functions exposed to the programmers to maintain simplicity. We also want the performance of the thread pool to be reasonably more efficient than normal thread-based version. Thus we create several test cases to compare the performances in different implementations. We mainly focused on testing the performance difference in thread-based implementation and Goroutine-based version for certain types of jobs. Finally, we simply want to learn how to implement a thread pool and explore the drawbacks along with possible improvmetns.\\

@@

\begin{itemize}
	\item \begin{lstlisting}
void thread_pool_init(int workers, 
		      int mutex_flag);
	\end{lstlisting}
	This function initialize a new thread pool. The \textbf{workers} attribute specifies the number of threads that can exist in a thread pool. If $0$ is passd in, the number of threads will be initialized to the number of core in the execution environment. The \textbf{mutex\_flag} argument let user decides whether to use mutex or spinlock as synchronization method. Note if the thread pool is initialized more than once, every invocation of this function will be discarded.\\
	\item 
	\begin{lstlisting}
bool thread_pool_add(task_func *func,
		     void* aux,
		     enum CallType);
	\end{lstlisting}
	This function adds tasks to the thread pool. The \textbf{func} attribute represents the function that is to be executed. The \textbf{aux} attribute indicates the parameter \textbf{func} takes in. the \textbf{CallType} specifies whether to use blocking or non-blocking mechanism for the thread pool. If a task is added without failure, $true$ is returned. Otherwise, $false$ will be returned.\\
	\item \begin{lstlisting}
void thread_pool_wait();
	\end{lstlisting}
	This function signifies all threads in thread pool that no more job will be added and the calling function waits for all tasks in the thread pool to finish.
\end{itemize}

\section{Implementation}
The low overhead of the thread pool is acheived with several design decisions. We will discuss the major implementations. A thread pool is first initialized by the main thread and the corresponding number of threads are initialized with their attributes. Each task is represented as a \textbf{struct task}. It is created by the calling function whenever the user sends the function to be executed to the thread pool. It is then distributed to the threads in the thread pool in a round-robin style to ensure even distribution. Each thread owns a privat task queue to store all its assigned tasks. Each queue is protected by either a mutex or a spinlock depending on users' specification. However, it is possible that a thread's task queue is empty when it is first initialized. It will waste CPU resource if such task is kept up and running without executing any tasks. Thus, we use semaphore to improve the efficiency of the thread pool. Thus we let the main thread invoking sema\_post on each thread once it finishes adding tasks to the thread's task queue to indicate the availability of tasks. However, the thread's task queue can be empty for other reasons. The most important one is that the thread finishes executing all assigned tasks. Thus, we implement work stealing algorithm to keep the idle thread busy by letting it trying to steal tasks from other threads by randomly choosing one. Since we don't want tasks stealing to create too much overhead as it tries to grab tasks from other threads, we simply let the idle thread using pthread\_mutex\_trylock() or pthread\_spin\_trylock(). If the stealing succeeds, the tasks is grabbed from the victim's task queue and executed. Otherwise, the steal will fail. If the steal fails too many times, that indicates many threads' task queue is already empty and the thread pool is getting close to running out of tasks. In this case, letting more threads trying to steal tasks will create unnecessary overhead, thus the thread will be blocked to prevent wasting resources. If eventually the threads run out of tasks, it will increase \textbf{wait\_sema} to let the main thread know it finishes its job. If all threads signify that they've finihsed their tasks, then thread pool is terminated.

\section{Analysis}
Below is the hardware we used to conduct experiements on the performance of the thread pool.
	\begin{table}[!hbt]
		% Center the table
	\begin{center}
		\caption{Hardware Sepcification}
		\label{tab:simParameters}
		% Table itself: here we have two columns which are centered and have lines to the left, right and in the middle: |c|c|
		\begin{tabular}{|c|c|}
			% To create a horizontal line, type \hline
			\hline
			% To end a column type &
			% For a linebreak type \\
			Processor & Intel® Core™ i7-6700 CPU \\
								&@3.40GHz $\times$ 8\\
			\hline
			RAM & $15.6$ GiB \\
			\hline
			OS & Ubuntu 16.04.4 LTS 64-bit\\
			\hline
		\end{tabular}
		\end{center}
	\end{table}

@@

\section{Conclusion}

	% If you have questions about how to write mathematical formulas in LaTeX, please read a LaTeX book or the 'Not So Short Introduction to LaTeX': tobi.oetiker.ch/lshort/lshort.pdf

	% This is how you include a eps figure in your document. LaTeX only accepts EPS or TIFF files.
	\begin{figure}[!hbt]
		% Center the figure.
		\begin{center}
		% Include the eps file, scale it such that it's width equals the column width. You can also put width=8cm for example...
		\includegraphics[width=\columnwidth]{plot_tf}
		% Create a subtitle for the figure.
		\caption{Simulation results on the AWGN channel. Average throughput $k/n$ vs $E_s/N_0$.}
		% Define the label of the figure. It's good to use 'fig:title', so you know that the label belongs to a figure.
		\label{fig:tf_plot}
		\end{center}
	\end{figure}

>>>>

% Define document title and author
	\title{JEThread: A Thread Pool Implementation in C}
	\author{Jialin Li and Edward Hu}
	\maketitle

% Write abstract here
\begin{abstract}
	JEThread is a Thread Pool written in C. It provides an easier interface and good performance for concurrent programming in C. JEThread achieves good performance across a mixture of workloads and automatically load balance through techniques like work stealing.
\end{abstract}

% Each section begins with a \section{title} command
\section{Introduction}
	% \PARstart{}{} creates a tall first letter for this first paragraph
	\PARstart{T}{hread} is a basic unit of CPU utilization, it consists of program counter, register set, and its own stack. Compared to processes, threads are much more lightweight since virtual memory space remains the same during a thread switch, which is not true for process switch. However, even threads sometimes pose overhead that will affect the performance of the program. Initializing and cleaning threads can pose a certain amount of overhead. For a concurrent application that creates and destroys a large number of threads that each only runs for a short time, the thread management mechanism can cause unnecessary resource waste. Thread pool can be useful in this type of scenario as it reduces the complexity of thread management and the overhead involved in thread creation and cleaning.

% Main Part
\section{Background}	
	Without a thread pool, the common techniques used to do concurrent programming in C are to spawn a thread for each task, if task is long running, or to create a fixed number of worker threads and statically partition the work among the workers. Although both ways are feasible and provide control freedom to programmers, they can also be unnecessary burdens and require programmers to have a great deal of knowledge of both the workloads they are running and the type of thread programming model that  would result in good performance. Newer languages like Golang deals with this problem by providing Goroutines which are extremely lightweight user threads, as well as other concurrent native objects to better support concurrent programming. These features have brought Golang many users. For example, container engines like Docker and LXD, container orchestration tools like Kubernetes are all written in Golang. While it is hard to fully integrate runtime concurrent programming support in C, the thread pool abstraction provides a good middle ground.

\section{Motivation}
	Our primary goal is to implement a thread pool that is easy to use and flexible with different workloads. We minimizes the number of functions in the thread pool interface to keep it narrow yet fully functional. We also make a distinction between blocking and nonblocking tasks, and adopt techniques like work stealing to support various workloads.
	We also want the thread pool to be reasonably more performant than previously mentioned pthread-based approaches. Thus we create several test cases to compare the performance in different implementations. Finally, we simply want to learn the inner-working of an efficient thread pool and explore the drawbacks along with possible improvements.\\

\begin{itemize}
	\item \begin{lstlisting}
void thread_pool_init(int workers, 
		      int mutex_flag);
	\end{lstlisting}
	This function initializes a new thread pool. The \textbf{workers} attribute specifies the number of threads that can exist in a thread pool. If $0$ is passd in, the number of threads will be initialized to the number of core in the execution environment. The \textbf{mutex\_flag} argument let user decides whether to use mutex or spinlock as synchronization method. Note if the thread pool is initialized more than once, every invocation of this function will be discarded.\\
	\item 
	\begin{lstlisting}
bool thread_pool_add(task_func *func,
		     void* aux,
		     enum CallType);
	\end{lstlisting}
	This function adds tasks to the thread pool. The \textbf{func} attribute represents the function that is to be executed. The \textbf{aux} attribute indicates the parameter \textbf{func} takes in. the \textbf{CallType} specifies whether to use blocking or non-blocking mechanism for the thread pool. If a task is added without failure, $true$ is returned. Otherwise, $false$ will be returned.\\
	\item \begin{lstlisting}
void thread_pool_wait();
	\end{lstlisting}
	This function blocks until all threads in thread pool have finished all the existing jobs.
		\end{itemize}

\section{Implementation}
The low overhead of the thread pool is achieved with several design decisions. We will discuss the major implementations. 
\subsection{Task Assignment}
Each task is represented as a \textbf{struct task}. It is created by the calling function whenever the user sends the function to be executed to the thread pool and then distributed to the threads in the thread pool in a round-robin style to ensure even distribution. We keep track of the current worker that we are assigning to, and advance it after we've assigned $N$ tasks to it. This is to reduce the overhead of waking up workers when the number of tasks is few, and to try batching tasks and provide better tasks locality. This is based on the hypothesis that tasks created close in time are likely to reference similar memory regions. Therefore, we hope to keep them on the same worker thread. Even when this hypothesis does not hold, it does not hurt JEThread's performance. When a job is assigned to a worker, the main thread grabs the lock(spinlock suggested) to the worker's private queue and add it to the queue.

In addition to batch assigning jobs, we also treat blocking and nonblocking tasks differently. This is necessary because JEThread by default starts with $N$ threads, where $N$ is the number of processors, and if all workers threads are executing some blocking tasks, then nonblocking tasks will not be able to execute until much later. It becomes even worse when majority of tasks are blocking tasks. In such case, the fixed number of workers would be blocked constantly. To solve this problem, we use an approach similar to Golang's parking threads, where we create a new thread to handle blocking tasks and only use our default set of worker threads to handle nonblocking tasks. In order to implement this, we need to identify if a task is blocking or nonblocking. This is difficult and inefficient to do at runtime, but can be done much easier in compile time. In our implementation, we rely on user to specify the CallType when adding a task. This is not fundamentally necessary - we could have modified GCC compiler or inspected task at preprocessing time to accomplish the same thing, but due to the time constraint we exposed this option to users. We believe that users would try their best to provide accurate CallType because they want better performance. Even if they accidentally misidentified CallType, this would not effect the correctness of JEThread.

\subsection{Internal Synchronization}
Each thread owns a privat task queue to store all its assigned tasks. Each queue is protected by either a mutex or a spinlock depending on users' specification. However, it is possible that a thread's task queue is empty when it is first initialized. It will waste CPU resource if such task is kept up and running without executing any tasks. Thus, we use semaphore to improve the efficiency of the thread pool. Thus we let the main thread invoking sema\_post on each thread once it finishes adding tasks to the thread's task queue to indicate the availability of tasks. However, the thread's task queue can be empty for other reasons. The most important one is that the thread finishes executing all assigned tasks. Thus, we implement work stealing algorithm to keep the idle thread busy by letting it trying to steal tasks from other threads by randomly choosing one. Since we don't want tasks stealing to create too much overhead as it tries to grab tasks from other threads, we simply let the idle thread using pthread\_mutex\_trylock() or pthread\_spin\_trylock(). If the stealing succeeds, the tasks is grabbed from the victim's task queue and executed. Otherwise, the steal will fail. If the steal fails too many times, that indicates many threads' task queue is already empty and the thread pool is getting close to running out of tasks. In this case, letting more threads trying to steal tasks will create unnecessary overhead, thus the thread will be blocked to prevent wasting resources. If eventually the threads run out of tasks, it will increase \textbf{wait\_sema} to let the main thread know it finishes its job. If all threads signify that they've finihsed their tasks, then thread pool is terminated.

\section{Evaluation}
Below is the hardware we used to conduct experiements on the performance of the thread pool.
	\begin{table}[!hbt]
		% Center the table
	\begin{center}
		\caption{Hardware Sepcification}
		\label{tab:simParameters}
		% Table itself: here we have two columns which are centered and have lines to the left, right and in the middle: |c|c|
		\begin{tabular}{|c|c|}
			% To create a horizontal line, type \hline
			\hline
			% To end a column type &
			% For a linebreak type \\
			Processor & Intel® Core™ i7-6700 CPU \\
								&@3.40GHz $\times$ 8\\
			\hline
			RAM & $15.6$ GiB \\
			\hline
			OS & Ubuntu 16.04.4 LTS 64-bit\\
			\hline
		\end{tabular}
		\end{center}
	\end{table}

@@

@@

\section{Conclusion}
	Write something here lol
